---
title: "Getting started with pollmap"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with pollmap}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

`pollmap` maps wild-bee pollination in agricultural landscapes. It brings the
distance-decay ("Lonsdorf") family of land-use models, a species-distribution
pathway, and pollinator-dependence demand mapping into one reproducible,
raster-native framework built on `terra`. This vignette shows the main
workflows; every function has its own help page with references.

## Installation

```{r}
# install.packages("remotes")
remotes::install_github("ehsanrahimi666/pollmap")
library(pollmap)
library(terra)
```

## 1. Land-use supply models

All land-use models share one foraging engine (`poll_kernel()`) and take a
land-cover raster plus parameter tables.

### The original Lonsdorf (2009) model

```{r}
lulc <- rast(nrows = 40, ncols = 40, xmin = 0, xmax = 4000, ymin = 0, ymax = 4000)
values(lulc) <- 1L; lulc[, 1:5] <- 2L          # farmland with a forest strip

lc <- poll_landcover(data.frame(lucode  = c(1L, 2L),
                                nesting = c(0.05, 1.0),
                                floral  = c(0.6, 0.2)))
g  <- poll_guild(alpha = 500, name = "wild_bee")

m  <- poll_lonsdorf(lulc, lc, g)
plot(m)
```

### The seasonal, guild-based InVEST model

`poll_kennedy()` uses the full InVEST schema (substrates x seasons x guilds).
Bundled default tables are provided for CORINE land cover and Swiss bees:

```{r}
bio    <- poll_biophysical(read.csv(system.file("extdata",
              "clc_biophysical_table.csv", package = "pollmap")))
guilds <- poll_guilds(read.csv(system.file("extdata",
              "swiss_guild_table.csv", package = "pollmap")))
mk <- poll_kennedy(clc_raster, bio, guilds)     # clc_raster = your CORINE raster
```

`poll_modified()` (PollMap) adds quality-weighted, flight-range foraging;
`poll_estimap()` adds abiotic activity modifiers; `poll_mce()` integrates extra
criteria by weighted linear combination (optionally with AHP weights).

## 2. Species-distribution supply

Fit occurrence models, project to suitability, and couple to a community supply
surface. The self-contained algorithms are `glm`/`gam`/`rf`; `flexsdm` can be
used for a fuller workflow.

```{r}
predictors <- rast(list.files("bioclim", full.names = TRUE))   # your predictors
occ <- read.csv("Bombus_terrestris.csv")                       # Species, Longitude, Latitude

fit <- poll_sdm(occ, predictors, method = c("glm", "rf"))
sui <- poll_sdm_project(fit, predictors)
supply <- poll_sdm_supply(list(sui_sp1, sui_sp2, sui_sp3))      # per-species maps
```

## 3. Demand and supply-demand balance

```{r}
dep    <- poll_crop_dependence(read.csv(system.file("extdata",
              "clms_crop_dependence.csv", package = "pollmap")),
              crop = "crop", dependence = "dependence")
demand <- poll_demand(crop_raster, dep)
bal    <- poll_balance(supply, demand)          # balance + deficit layers
plot(bal)
```

## 4. Analysis utilities

These operate on any model output.

```{r}
# land-use-change scenario: convert forest (2) to farmland (1)
sc <- poll_scenario(poll_lonsdorf, lulc, lc, g, from = 2L, to = 1L)
plot(sc$delta)

# uncertainty from resource scores (e.g. Koh et al. 2016 Beta priors)
draw <- beta_landcover(koh_priors)              # priors with shape1/shape2 columns
unc  <- poll_sensitivity(run = function() poll_lonsdorf(lulc, draw(), g), n = 100)

# compare several models on identical inputs
cmp <- poll_compare(lonsdorf = m$map, kennedy = mk$map[["supply"]])

# validate against field data
val <- poll_validate(m, observed_points, coords = c("x", "y"))

# classify into ordinal classes with an area table
cl  <- poll_classify(m, n = 5)
```

## 5. Neutral landscapes

For testing and structure experiments, without external data:

```{r}
land <- poll_simulate(nrow = 100, ncol = 100, p_habitat = 0.3, autocorr = 3)
frag <- poll_fragment(p_seq = c(0.1, 0.3, 0.5), autocorr_seq = c(0, 3))
mets <- poll_metrics(m, lulc)                   # per-class supply (+ landscapemetrics)
```

## Scope and caveats

`pollmap` produces relative indices of supply, abundance and demand for
ecological analysis, not economic valuation. The land-use models inherit the
assumptions of their sources: abundance is a proxy for pollination, floral
resources are not depleted, and the default parameter tables are
literature-informed expert values intended to be reviewed and overridden. Use
`poll_sensitivity()` and `poll_validate()` to quantify and check these.
```
